Efficient forwards-time simulation with ancestral recombination graphs

### 2017-10-23 forwards_paper.tex
Added asy source for rough version of diagram.
Also short discussion of sites and mutations.

    Coalescent simulations are very helpful but require random mating and neutrality.
    For continuous space, polygenic selection, or detailed dissection of life history,
    we must use forwards-time, individual-based simulation.
    These are much slower due in part to carrying around neutral genotypes irrelevant to the process.
    Here we show how to efficiently produce and store the entire history of ancestry and recombination
    (the ARG) from an individual-based simulation,
    on which neutral mutations can be placed afterwards.
    This has the promise of making large-scale, whole-genome simulations with realistic geography and selection finally possible.
\end{abstract}

@@

In so doing we record the \emph{population pedigree} --
the entire history of parent-offspring relationships of an entire population going back to a remote time --
as well as information encoding the genetic outcomes of each ancestral meiosis --
who inherited which parts of which parental chromosomes.
This embellished graph is known as the \emph{ancestral recombination graph}, or {ARG} \citep{griffiths}.
Combined with ancestral genotypes and the origins of new mutations,
it completely specifies the genomic sequence of any individual in the population at any time.
However, much less than the entire ARG is needed to specify relationships
between any given set of samples --
only those portions of it from which those samples have actually inherited,
back to their most recent common ancestors.
The assumptions of coalescent theory
% -- random mating and neutrality --
imply that this random set describes a Markov process looked at \emph{backwards} in time,
and can hence be simulated without reference to the unnecessary remainder
\citet{griffithsmarjoram,msprime}.
This stochastic process is also called the ``ARG'',
but we reserve this name for the embellished graph itself.
% Although this point was in principle known since \citet{griffithsmarjoram},
% only recent algorithmic advances made it possible to actually simulate the process correctly,
% without approximation, across whole chromosomes \citep{msprime}.
\jk{I find this confusing I'm afraid. I agree we should mention the ARG but I think only in passing.
I have never found a situation in which an ARG made things easier to understand!}

@@

\begin{figure}
    \begin{center}
        \includegraphics[width=0.6\textwidth]{example_tree_sequence}
    \end{center}
    \caption{
        An example tree sequence with three samples over a chromosome of length 10. In the
        left-most panel we show the tree sequence pictorially in two different ways:
        in the top are the tree topologies and the bottom shows the spatial extent of the
        edges that define these topologies. The right-hand panels show the specific encoding
        of this tree sequence in the four tables (nodes, edges, sites and mutations) defined
        by \msprime.
        \label{fig:example_tree_sequence}
    }
\end{figure}

@@

\jk{Continue along these lines with sites and mutations after updating the figure}

@@

\subsection*{Data structures:}

\begin{figure}
    \begin{center}
        \includegraphics[width=0.6\textwidth]{example_tree_seq}
    \end{center}
    \caption{
        A pictorial representation of the ARG
        relating three samples to each other
        over a chromosome of length 1.0.
        Mutations shown in the example tables of the text
        are marked with red stars.
        \label{fig:ex_tree_seq}
    }
\end{figure}

%%%%%%
\subsection*{Recording the ARG in forwards time}

@@

\begin{figure}
    \begin{center}
        \includegraphics{method_diagram}
    \end{center}
    \caption{
        A simple example of the method.
        \textbf{Top:} the ARG shown on the left relates ten haploid individuals to each other.
        It is recorded, in forwards time,
        in 10 node records (one for each individual)
        and 14 edgeset records (one for each distinctly inherited segment).
        Blue numbers denote crossing over locations in each meiosis.
        The individuals $B$, $C$, and $D$ inherit clonally from $A$
        to ensure rootnedness of resulting tree.
        \textbf{Center:} the five distinct trees relating all individuals to each other
        found across the chromosome (blue numbers denote locations on the chromosome).
        Labels after simplification are shown in red.
        \textbf{Bottom:} tables recording the tree sequence after simplification
        with nodes $I$ and $J$ as samples.
        The mapping from labels in the forwards time simulation to nodes in the tree sequence
        is shown in red,
        which allows additional records to be added as the simulation progresses.
        \label{fig:method_diagram}
    }
\end{figure}

@@

\section{More general method for recording the ARG}

>>>>

    Coalescent simulations are very helpful but require random mating and neutrality.
    For continuous space, polygenic selection, or detailed dissection of life history,
    we must use forwards-time, individual-based simulation.
    These are much slower due in part to carrying around neutral genotypes irrelevant to the process.
    Here we show how to efficiently produce and store the entire history of ancestry and recombination
    from an individual-based simulation,
    on which neutral mutations can be placed afterwards.
    This has the promise of making large-scale, whole-genome simulations with realistic geography and selection finally possible.
\end{abstract}

@@

In so doing we record the \emph{population pedigree} --
the entire history of parent-offspring relationships of an entire population going back to a remote time --
as well as information encoding the genetic outcomes of each ancestral meiosis --
who inherited which parts of which parental chromosomes.
This embellished pedigree contains all the information necessary
to construct the genealogical tree that relates each individual to each other
at each position on the genome, i.e., the \emph{tree sequence}.
Combined with ancestral genotypes and the origins of new mutations,
it also completely specifies the genomic sequence of any individual in the population at any time.
This is much more than we need to know, however,
so we discard all information irrelevant to the genetic history
of the \emph{sampled} individuals,
which results in considerable savings.

Another way of representing this same information
is known as the \emph{ancestral recombination graph}, or {ARG} \citep{griffiths1997ancestral},
which has been the subject of substantial study
under the assumptions of coalescent theory \citep{wiuf1997number,wiuf1999ancestry,marjoram2006coalescent,wilton2015smc}.

@@

\begin{figure}
    \begin{center}
        \includegraphics[width=\textwidth]{example_tree_sequence}
    \end{center}
    \caption{
        An example tree sequence with three samples over a chromosome of length 10. In the
        left-most panel we show the tree sequence pictorially in two different ways:
        in the top are the tree topologies and the bottom shows the spatial extent of the
        edges that define these topologies. The right-hand panels show the specific encoding
        of this tree sequence in the four tables (nodes, edges, sites and mutations) defined
        by \msprime.
        \jk{The figure needs to be rescaled a bit to give more space to the trees. We could also use a
        time scale next to the trees.}
        \label{fig:example_tree_sequence}
    }
\end{figure}

@@

Given the topology defined by the nodes and edges, \emph{sites} and \emph{mutations}
encode the sequence information for each sample in an efficient way. Each site
is associated with a position on the genome and an ancestral state. For example,
in Figure~\ref{fig:example_tree_sequence} we have two sites, one at position
2 and ancestral state `A' and the other at position 7 with ancestral state `G'. If
no mutations occur, all samples inherit the ancestral state at a given site.
A mutation occurs over a specific node at a given site, and results in a specific
derived state. Thus, all samples below the mutation node in the tree will inherit
this state (unless further mutations are encountered). Three mutations are shown
in Figure~\ref{fig:example_tree_sequence}, illustrated by red hexagons. The first
mutation occurs at site zero (which is in the left-hand tree), which is a simple
mutation resulting in node $2$ inheriting the state 'T'. The second side (in the
right hand tree) has two mutations: one occuring over node $3$ changing the state to
`C' and a back mutation over node $1$ changing the state to `G'.
\jk{Not sure if I'm labouring the point here: this should be reasonably obvious.
Perhaps if we added the state changes to the trees, and the states of the leaves
below the samples, we could get rid of all the example discussion here?}

@@

%%%%%%
\subsection*{Recording the pedigree in forwards time}

@@

\begin{figure}
    \begin{center}
        \includegraphics{method_diagram}
    \end{center}
    \caption{
        A simple example of the method.
        \textbf{Top:} the diagram on the left relates ten haploid individuals to each other.
        It is recorded, in forwards time,
        in 10 node records (one for each individual)
        and 14 edgeset records (one for each distinctly inherited segment).
        Blue numbers denote crossing over locations in each meiosis.
        The individuals $B$, $C$, and $D$ inherit clonally from $A$
        to ensure rootnedness of resulting tree.
        \textbf{Center:} the five distinct trees relating all individuals to each other
        found across the chromosome (blue numbers denote locations on the chromosome).
        Labels after simplification are shown in red.
        \textbf{Bottom:} tables recording the tree sequence after simplification
        with nodes $I$ and $J$ as samples.
        The mapping from labels in the forwards time simulation to nodes in the tree sequence
        is shown in red,
        which allows additional records to be added as the simulation progresses.
        \label{fig:method_diagram}
    }
\end{figure}

@@

\subsection*{Data structures:}
\jk{Moved from above.}

\begin{figure}
    \begin{center}
        \includegraphics[width=0.6\textwidth]{example_tree_seq}
    \end{center}
    \caption{
        A pictorial representation of a tree sequence
        relating three samples to each other
        over a chromosome of length 1.0.
        Mutations shown in the example tables of the text
        are marked with red stars.
        \label{fig:ex_tree_seq}
    }
\end{figure}

\section{More general method for recording the pedigree}
